---
title: "ETV - Task - US States County Population"
format: html
editor: visual
---

## Description

Extracting county population data from Wikipedia provides a practical approach to measuring the size of the United States in an **informal census**, offering insights into **population distribution and density**. By leveraging web scraping techniques with R, we can systematically collect population data for each county across all 50 states. This process involves identifying structured tables on Wikipedia, extracting relevant columns—such as county names and population figures—and standardizing the data for analysis. Since county population figures are regularly updated, this method provides a **near real-time snapshot** of demographic patterns without relying on official census releases.

The purpose of this informal census is to assess **population density trends**, compare regional distributions, and explore variations in urban versus rural areas. By aggregating county-level data, we can better understand how population is concentrated across states, which is useful for studies in **resource allocation, infrastructure planning, and socioeconomic analysis**. While this approach does not replace official census data, it serves as a valuable **exploratory tool** for estimating and analyzing demographic shifts across the United States.

Data Ethics:

When extracting county population data from Wikipedia, it is essential to consider the **ethical implications** surrounding web scraping, data usage, and interpretation. While Wikipedia is an open-source platform that allows public access to information, web scraping should be conducted **responsibly** to avoid excessive server requests that could disrupt website performance. Adhering to respectful data collection practices—such as adding time delays between requests and checking the website’s **robots.txt** file—ensures that data extraction aligns with ethical guidelines and does not impose an undue burden on the platform. Additionally, since Wikipedia is a **crowdsourced** knowledge repository, it is crucial to verify the accuracy of extracted data by cross-referencing with official government sources like the **U.S. Census Bureau**.

Beyond ethical data collection, the **use and interpretation of extracted data** also carry important responsibilities. Population data can influence **policy discussions, economic decisions, and social analyses**, making it essential to avoid **misrepresentation or misuse** of the information. Researchers and analysts should be transparent about the **limitations** of web-scraped data, especially considering that Wikipedia figures may not always be up-to-date or methodologically rigorous. Furthermore, when analyzing population density or demographic trends, it is critical to **avoid biases** that could lead to misleading conclusions about regional disparities or socioeconomic conditions. Ethical data practices require that insights drawn from informal sources like Wikipedia be presented **with context, transparency, and acknowledgment of potential inaccuracies**, ensuring responsible and meaningful contributions to public discourse.

## Tasks:

To analyze the distribution of county population sizes across the United States, create a histogram using ggplot2, applying a log10 transformation to the population data. Given that county populations vary dramatically—from low-density rural areas to highly populated urban centers—log transformation helps normalize the data, making patterns more interpretable. The histogram should reveal how counties are distributed across different population scales, highlighting whether the data is skewed, if most counties have smaller populations, or if there are outliers with significantly higher populations. This visualization offers insights into population density trends, helping to distinguish between predominantly rural and highly urbanized counties.

Run the following code to install the `devtools` and `purrr` packages:

```{r}
# install.packages("devtools")
# install.packages("purrr")
```

Then, use **devtools** to install the **htmltab** package from GitHub:

```{r}
# devtools::install_github("crubba/htmltab", force = TRUE)
```

The **htmltab** package provides a function that simplifies extracting HTML tables from Wikipedia pages.

#### Load Libraries

\[Clear this - Begin\]

```{r}
library(htmltab)  ## Data Extraction
library(dplyr)    ## Data Transformation
library(stringr)  ## Data Transformation
library(tidyr)    ## Data Transformation
library(purrr)    ## Data Transformation
library(ggplot2)  ## Data Visualization
```

\[Clear this - End\]

### Extraction

\[Delete this - Begin\]

```{r}
url <- "https://en.wikipedia.org/wiki/List_of_counties_in_California" ## Note 1: Look at url and draw attention images are in table
```

\[Delete this - End\]

```{r}
# url <- ""
```

\[Delete this - Begin\]

```{r}
ca_df <- htmltab(doc = url, which = 1) ## Note 2: which allows you to select the table you want to extract
ca_df <- htmltab(doc = url, which = 2) ## Note 3: the warning mesaage is not an error and what it says
ca_df <- htmltab(doc = url, which = 2, rm_nodata_cols = FALSE) ## Note 4: rm_nodata_cols = FALSE keeps columns with no data
```

\[Delete this - End\]

```{r}
## Use the `htmltab` function to extract the data from the Wikipedia page. 
```

**Q1.** What is the primary purpose of the `htmltab` function? What does the `which` argument do?

\[Delete this - Begin\]

Solution: The `htmltab` function is used to extract HTML tables from Wikipedia pages. The `which` argument allows you to specify which table to extract from the page based on its position in the HTML code.

\[Delete this - End\]

### Transformation

Let's Focus on County, Population, and Area

\[Delete this - Begin\]

```{r}
mod_1_ca_df <- ca_df %>% 
  select(County, `Population(2023)`, Area) 
```

\[Delete this - End\]

```{r}
# mod_1_ca_df <- ca_df %>% 
 
```

**Q2.** What do you notice when you run `glimpse(mod_1_ca_df)` regarding the `Population(2023)` column?

\[Delete this - Begin\]

Solution: The `Population(2023)` column is a character column. We need to convert it to a numeric column. But before we do that, we need to remove the commas from the numbers.

\[Delete this - End\]

**Now, let's remove the commas from the `Population(2023)` column.**

\[Delete this - Begin\]

```{r}
mod_2_ca_df <- mod_1_ca_df %>% 
  mutate(`Population(2023)` = str_replace_all(`Population(2023)`, ",", ""))
```

\[Delete this - End\]

```{r}
# mod_2_ca_df <- mod_1_ca_df %>% 
#  mutate(`Population(2023)` = _______________(`Population(2023)`, ",", ""))
```

**Q3.** What do we still need to do regarding the `Population(2023)` column?

\[Delete this - Begin\]

Solution: We need to convert the `Population(2023)` column to a numeric column.

\[Delete this - End\]

**Now, let's convert the `Population(2023)` column to a numeric column.**

\[Delete this - Begin\]

```{r}
mod_3_ca_df <- mod_2_ca_df %>% 
  mutate(`Population(2023)` = as.numeric(`Population(2023)`))  

```

\[Delete this - End\]

```{r}
# mod_3_ca_df <- mod_2_ca_df %>% 
#   mutate(`Population(2023)` = ___________(`Population(2023)`))  

```

I do not like the column names being upper case and the parentheses in `Population(2023)` is distracting. Let's fix that.

\[Delete this - Begin\]

```{r}
mod_4_ca_df <- mod_3_ca_df %>% 
  rename(county = County, population = `Population(2023)`, area = Area) 
```

\[Delete this - End\]

```{r}
# mod_4_ca_df <- mod_3_ca_df %>% 
#  
```

**Q4.** Now run `glimpse(mod_4_ca_df)` to see the changes. When do you notice about area column?

\[Delete this - Begin\]

Solution: The `Area` column is a character column. **Also it contains two values: square miles and square kilometers.** We need to separate the two values into two separate columns.

\[Delete this - End\]

**Now, let's separate the `Area` column into `square_miles` and `square_kilometers` columns.**

\[Delete this - Begin\]

```{r}
mod_5_ca_df <- mod_4_ca_df %>% 
  separate(area, into = c("square_miles", "square_kilometers"), sep = "sq") 

```

\[Delete this - End\]

```{r}
# mod_5_ca_df <- mod_4_ca_df %>% 
#
```

Well that helped but the square miles and square kilometers columns are still character columns.

We also have some unwanted text in both columns. We need to convert them to numeric columns.

This is quite a bit of work.

Let's do it in steps.

**Steps:**

\[Delete this - Begin\]

1.  Replace all commas with empty strings
2.  Trim any leading or trailing white space
3.  Remove any unwanted text such as "mi", "km", ",", "(", and ")"
4.  Convert the columns to numeric

\[Delete this - End\]

\[Delete this - Begin\]

```{r}
mod_6_ca_df <- mod_5_ca_df %>% 
  mutate(square_miles = as.numeric(str_trim(str_replace_all(square_miles, ",","") )),
         square_kilometers = as.numeric(str_trim(str_replace_all(square_kilometers, "mi|km|,|\\)|\\(","") )), 
         state = "CA")
```

\[Delete this - End\]

```{r}
# mod_6_ca_df <- mod_5_ca_df %>%
```

Now do that for all 50 states in the United States with "one" line of code.

#### **Learning Moment 1 - Conditional Statements in R (Digression)**

1.  **if-else Statements**

    -   *Example:*\

    ```{r}
    x <- 10
    if (x > 5) {
     print("x is greater than 5")
    } else {
     print("x is 5 or less")
    }
    ```

2.  **Nested if-else Statements**

    -   *Example:*

```{r}
x <- 10
if (x > 10) {
 print("Greater than 10")
} else if (x == 10) {
 print("Exactly 10")
} else {
 print("Less than 10")
}
```

3.  **Logical Operators in Conditionals**

    -   *Example:*\

    ```{r}
    x <- 7
    if (x > 5 & x < 10) {
     print("x is between 5 and 10")
    }
    ```

4.  **Vectorized ifelse() Function**

    -   *Example:*\

    ```{r}
    numbers <- c(1, 5, 10, 15)
    result <- ifelse(numbers > 5, "Greater than 5", "5 or less")
    print(result)
    ```

5.  **switch() Function for Multiple Cases**

    -   *Example:*\

    ```{r}
    day <- "Monday"
    activity <- switch(day,
     "Monday" = "Start the week",
     "Friday" = "Prepare for the weekend",
     "Sunday" = "Relax",
     "Unknown day"
    )
    print(activity)
    ```

**Questions:**

**Q5.** What is the difference between using `if-else` and `ifelse()` in R?

\[Delete this - Begin\]

Solution: The `if-else` statement is used for conditional execution of code based on a single condition, while `ifelse()` is a vectorized function that applies a condition to each element of a vector.

\[Delete this - End\]

**Q6.** How does the `switch()` function differ from using multiple `if-else` statements?

\[Delete this - Begin\]

Solution: The `if-else` statement is used for conditional execution of code based on a single condition, while `ifelse()` is a vectorized function that applies a condition to each element of a vector. The `switch()` function is used to select one of several alternatives based on a single value.

\[Delete this - End\]

**Q7.** What is the purpose of conditional statements in data analysis?

\[Delete this - Begin\]

Solution: Conditional statements are used to control the flow of a program based on specified conditions. They allow you to execute different code blocks based on whether a condition is true or false.

\[Delete this - End\]

------------------------------------------------------------------------

#### **Learning Moment 2 - Iterations in R (Digression)**

1.  **for Loops**

    -   *Example:*\

    ```{r}
    for (i in 1:5) {
     print(paste("Iteration:", i))
    }
    ```

2.  **while Loops**

    -   *Example:*\

    ```{r}
    x <- 1
    while (x <= 5) {
     print(x)
     x <- x + 1
    }
    ```

3.  **repeat Loops with break Statements**

    -   *Example:*\

    ```{r}
    x <- 1
    repeat {
     print(x)
     x <- x + 1
     if (x > 5) break
    }
    ```

4.  **Using next to Skip Iterations**

    -   *Example:*\

    ```{r}
    for (i in 1:5) {
     if (i == 3) next
     print(i)
    }
    ```

**Questions:**

**Q8.** How does `repeat` differ from `while`, and when would you use it?

\[Delete this - Begin\]

Solution: The `repeat` loop is an infinite loop that continues executing until a `break` statement is encountered. It is useful when you need to repeatedly execute a block of code without a specific condition for termination.

\[Delete this - End\]

------------------------------------------------------------------------

#### **Learning Moment 3 - Functions in R (Digression)**

1.  **Basic Function Definition**

    -   *Example:*\

    ```{r}
    add_numbers <- function(a, b) {
     return(a + b)
    }
    print(add_numbers(3, 5))
    ```

2.  **Default Parameter Values**

    -   *Example:*\

    ```{r}

    greet <- function(name = "User") {
     return(paste("Hello,", name))
    }

    print(greet())

    print(greet("Alice"))
    ```

3.  **Returning Multiple Values (Lists)**

    -   *Example:*\

    ```{r}
    stats <- function(x) {
     list(mean = mean(x), median = median(x))
    }
    print(stats(c(1, 2, 3, 4, 5)))
    ```

4.  **Higher-Order Functions (Functions as Arguments)**

    -   *Example:*\

    ```{r}
    apply_function <- function(f, x) {
     return(f(x))
    }
    print(apply_function(sqrt, 16))
    ```

**Questions:**\
**Q9.** Create a function that takes two arguments, `a` and `b`, and returns the product of `a` and `b`. Allow the default value of `b` to be 3. Test it out when a = 4.

\[Delete this - Begin\]

Solution:

```{r}
product <- function(a, b = 3) {
 return(a * b)
}
print(product(4))
```

\[Delete this - End\]

------------------------------------------------------------------------

#### **Learning Moment 4 - Using purrr in R (Digression)**

1.  **Using map() for Iteration**
    -   *Example:*

\[Delete this - Begin\]

```{r}
numbers <- list(1, 2, 3, 4, 5)        ## Demonstration Change to vector
squared <- map(numbers, ~ .x^2)       ## Demonstration Change to map_dbl
print(squared)
```

\[Delete this - End\]

```{r}
# numbers <- list(1, 2, 3, 4, 5)        ## Demonstration Change to vector
# squared <- _____________________      ## Demonstration Change to map_dbl
# print(squared)
```

**Q10.** What is the purpose of `map()` in the purrr package?

\[Delete this - Begin\]

Solution: The `map()` function in the purrr package is used to iterate over elements of a list or vector and apply a function to each element, returning the results in a list.

\[Delete this - End\]

**Q11.** What does the `~` symbol represent in the `map()` function?

\[Delete this - Begin\]

Solution: The `~` symbol is used to define an anonymous function (lambda function) in R. It allows you to create a function on the fly without explicitly defining it.

\[Delete this - End\]

2.  **map2() for Pairwise Iteration**
    -   *Example:*

\[Delete this - Begin\]

```{r}
result <- map2(c(1, 2, 3), c(4, 5, 6), ~ .x + .y)
print(result)
```

\[Delete this - End\]

```{r}
# result <- _____(c(1, 2, 3), c(4, 5, 6), ________)
# print(result)
```

**Q12.** What is the purpose of `map2()` in the purrr package?

\[Delete this - Begin\]

Solution: The `map2()` function in the purrr package is used to iterate over two input vectors in parallel and apply a function to corresponding elements.

\[Delete this - End\]

**Q13.** What does the "." in the `map2()` function?

\[Delete this - Begin\]

Solution: The "." symbol is used to refer to the elements of the input vectors being iterated over in the `map2()` function. It allows you to access and operate on the elements within the function definition.

\[Delete this - End\]

3.  **pmap() for Iterating Over Multiple Inputs**
    -   *Example:*

\[Delete this - Begin\]

```{r}
df <- data.frame(a = 1:3, b = 4:6, c = 7:9)
result <- pmap(df, sum)
print(result)
```

\[Delete this - End\]

```{r}
# df <- data.frame(a = 1:3, b = 4:6, c = 7:9)
# result <- ____(df, sum)
# print(result)
```

**Q14.** What is the purpose of `pmap()` in the purrr package?

\[Delete this - Begin\]

Solution: The `pmap()` function in the purrr package is used to iterate over multiple input vectors or data frames in parallel and apply a function to corresponding elements.

\[Delete this - End\]

**Questions:**

**Q15.** How does `map()` improve over traditional `for` loops in R?

\[Delete this - Begin\]

Solution: The `map()` function simplifies the process of iterating over elements of a list or vector by applying a function to each element and returning the results in a list. This is more concise and readable than using traditional `for` loops.

\[Delete this - End\]

**Done with digressions.**

Lets using what we learned to create a function that extracts and transforms data from Wikipedia pages for all states in the United States.

```{r}

url <- "https://en.wikipedia.org/wiki/List_of_counties_in_"
states_url <- str_c(url, state.name) %>% 
  str_replace_all(" ", "_") 

```

```{r}
get_wiki_data <- function(url, state_name) {
  
  ## Step 1: Define special states that require different table numbers
  special_states_tab_1 <- c("Minnesota","Ohio") %>% 
    str_c(sep = "|")
  
  special_states_tab_3 <- c("South Carolina","Arkansas","Georgia","Louisiana","New Mexico","Virginia","New York" ) %>% 
    str_c(sep = "|")
  
  if (state_name %in% special_states_tab_1) {
    table_number = 1
  } else if (state_name %in% special_states_tab_3) {
    table_number = 3
  } else {
    table_number = 2
  }
  
  ## Step 2: Extract data from Wikipedia page
  df <- htmltab(doc = url, 
                which = table_number,
                rm_nodata_cols = FALSE)
  
  ## Step 3: Rename columns and select relevant columns
  mod_1_df <- df %>% 
    rename_with(~ "county", any_of(c("Parish", "Borough","County"))) %>%
    rename_with(~ "population", any_of(c("Population", "Pop.", "Population(2023)", 
                                        "Pop. (2023)", "Pop.(2023)", "Population(2023 Estimate)"))) %>%
    rename_with(~ "area", contains("Area")) %>% 
    select(county, population, area) 
  
  
  ## Step 4: Remove commas from population column and convert to numeric
  mod_2_df <- mod_1_df %>%
    mutate( population = str_replace_all(population, ",", "")) %>%
    mutate(population = as.numeric(population)) 
  


  ## Step 4: Separate the area column into square miles and square kilometers
  mod_3_df <- mod_2_df %>% separate(area, into = c("square_miles", "square_kilometers"), sep = "sq") 
  
  
  
  ## Step 5: Convert to numeric and remove unwanted text and create state column
  mod_4_df <- mod_3_df %>% mutate(square_miles = as.numeric(str_trim(str_replace_all(square_miles, ",","") )),
         square_kilometers = as.numeric(str_trim(str_replace_all(square_kilometers, "mi|km|,|\\)|\\(","") )), 
         state = state_name) %>%
    
  return(mod_4_df)
}

```

```{r}
## Iterate over the `states_vec` and `states_url` vectors to extract and transform data for all U.S. states. 
## If an error occurs during extraction for a particular state, log the error message along with the state name 
## and continue processing the remaining states. Introduce a 5-second delay between iterations to avoid overloading the server. 
## Store any errors in a data frame for later review.

Initialize storage for state data and errors
state_dfs <- list()
error_log <- data.frame(state = character(), error_message = character(), stringsAsFactors = FALSE)

for (i in seq_along(state.name)) {
  tryCatch({
    state_dfs[[i]] <- get_wiki_data(states_url[i], state.name[i])  # Ensure correct variable for state names
  }, error = function(e) {
    message <- paste("Error extracting data for", state.name[i], ":", e$message)
    print(message)

    # Store error details in the error_log data frame
    error_log <<- bind_rows(error_log, data.frame(state = state.name[i], error_message = e$message, stringsAsFactors = FALSE))
  })
  print(paste("Data extracted for", state.name[i],"\n"))

  Sys.sleep(2)  # Introduce delay to prevent server overload
}


```

```{r}
state_info_dfs <- map2_df(states_url, state.name, get_wiki_data)
```

### Visualization

## Recap
